---
title: 'Module 05: One Sample Mean'
author: "Patrick J. Rosopa, Ph.D."
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: yes
    toc_depth: 3
  word_document:
    toc: yes
    toc_depth: 3
  html_document:
    toc: yes
    toc_depth: '3'
    df_print: paged
editor_options: 
  chunk_output_type: console
---

## Objective

-   Practice analyses involving a single sample mean using the contrived data on depression.

### Import data

Recall that there are **several** ways to load a dataset into `R`:

-   If the data is not too large, you can type the data into `R`

-   Importing data over the Internet

`dat <- read.csv('insert_link_to_file_here', header = T)`

-   Read the dataset from your computer

```{r}
# Download the data & ensure it is located in your project. In this example, the
# data file is located in a folder called Data in my current working directory.

library(haven)
depress <- read_sav("Data/1-One-sample t.sav")

str(depress)
```

Notice that the file is an *SPSS* data file. We cannot use `read.csv()` to import an *SPSS* data file. Above, we used a function in the `haven` package. When importing the data, notice that `read_sav()` created an object called a tibble. Think of a tibble as a data frame with some additional features. When using the `str()` function, we can see that there are 10 observations and only one variable in the tibble. The **depression** variable can be indexed using the \$.

### Examine the data

We can calculate descriptive statistics and examine a histogram of the outcome---*depression*.

```{r}
summary(depress)

mean(depress$depression)
sd(depress$depression)

hist(depress$depression, main = "", 
     xlab = "Depression",
     col = "slategray3",
     las = 1)
```

For continuous variables, we can also use `boxplot()` to get a sense of the center and the spread of the data. If you do not have too many different values of a continuous variable, a stem-and-leaf plot can also be useful to examine. Use `stem()` to examine the distribution of the scores. We can see that 6 occurs once. We can see that 35 occurs five times.

```{r}
stem(depress$depression)
```

### Conduct a two-tailed test on a single mean

With the `t.test()` function in base `R`, the dependent variable is the first argument. Use the optional `mu` argument to specify the value that we are testing against. By default, the test value is set to 0. However, for this problem, we specify 15. By default, a two-tailed test is conducted. A two-tailed test is a non-directional hypothesis test. We allow for the possibility that the null hypothesis can be false in either direction (i.e., $\mu < 15$ or $\mu > 15$).

```{r}
t.test(depress$depression, mu = 15)
```

We can see that the *t* random variable on 9 degrees of freedom is 3.4943. The *p* value is smaller than the conventional values for $\alpha$ like .05 and .01. Thus, we reject the null hypothesis. Note that the output also displays the alternative hypothesis which is that the "true mean is not equal to 15" ($\mu \ne 15$). In addition to the sample mean of 28.1, the 95% confidence interval is displayed. If some other confidence level is desired, not surprisingly, an optional argument can be used in the `t.test()` function. For example, if we want a 99% confidence interval, as expected, the estimated interval will be wider than that from a 95% confidence interval.

```{r}
t.test(depress$depression, mu = 15, conf.level = .99)
```

### Conduct a one-tailed test on a single mean

To test whether our sample comes from a population where the mean is *greater* than 15 ($\mu > 15$), we can use an optional argument. This is a one-tailed test where we believe that $\mu > 15$. If our directional hypothesis test is correct, then we would expect that our test statistic would be a positive number. Based on the very small *p* value less than $\alpha$, the population mean is, in fact, significantly greater than 15.

```{r}
t.test(depress$depression, mu = 15, alternative = "greater")
```

In the example above, we conducted a directional test with the rejection region in the right tail ($\mu > 15$). We also could have conducted a directional test with the rejection region in the left tail ($\mu < 15$). In this case, we would need to change the optional argument to alternative = "less."

Note: It is up to you as a researcher to decide whether to conduct a two-tailed test vs. a one-tailed test. If you conduct a one-tailed test, you should have strong theoretical support for the expected direction. In general, two-tailed tests appear to be more commonly used than one-tailed tests.

We can use `help("t.test")` to learn more about this function and the arguments that are required vs. optional. Instead of the `t.test()` function in base `R`, we can also use a function in the `lsr` package. Here, we conduct a two-tailed test.

```{r}
library(lsr)
oneSampleTTest(depress$depression, mu = 15)
```

This function produces much of the same output as the `t.test()` function. In addition, an estimate of Cohen's *d* is computed.
