---
title: "PSYC 8100: Exam 1"
author: "Alex Schlesener"
date: "2023-09-27"
output:
    pdf_document:
      toc: yes
      toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1: NASA-TLX
## 1a) Data Analyses (R)
### Load & Examine the dataset
```{r}
# Load the dataset
dat1 <- read.csv('Data/chemical.csv', header = T)

# Examine the data
str(dat1)
summary(dat1)

# Examine the means and standard deviations of each NASA-TLX dimension 
# (mental demand, physical demand, temporal demand, performance, effort, and frustration)
mean(dat1$mental)
mean(dat1$physical)
mean(dat1$temporal)
mean(dat1$performance)
mean(dat1$effort)
mean(dat1$frustration)

sd(dat1$mental)
sd(dat1$physical)
sd(dat1$temporal)
sd(dat1$performance)
sd(dat1$effort)
sd(dat1$frustration)
```
### Conduct a Test
Now we will conduct a t-test to compare the sample mean from mental demand (5.45) with the sample mean from previous publishing involving normal
workload (4.5).
```{r}
t.test(dat1$mental, mu = 4.5)
```
Now we will conduct a t-test to compare the sample mean from physical demand (4.646) with the sample mean from previous publishing involving normal
workload (4.5).
```{r}
t.test(dat1$physical, mu = 4.5)
```

Now we will conduct a t-test to compare the sample mean from temporal demand (5.60) with the sample mean from previous publishing involving normal
workload (4.5).
```{r}
t.test(dat1$temporal, mu = 4.5)
```

Now we will conduct a t-test to compare the sample mean from performance (5.48) with the sample mean from previous publishing involving normal
workload (4.5).
```{r}
t.test(dat1$performance, mu = 4.5)
```

Now we will conduct a t-test to compare the sample mean from effort (5.50) with the sample mean from previous publishing involving normal
workload (4.5).
```{r}
t.test(dat1$effort, mu = 4.5)
```

Now we will conduct a t-test to compare the sample mean from frustration (4.60) with the sample mean from previous publishing involving normal
workload (4.5).
```{r}
t.test(dat1$frustration, mu = 4.5)
```

### Assess Positive Association
Now, we will check for the possibility of positive association between mental, physical, and temporal demands.

```{r}
# Correlation between mental and physical
cor(dat1$mental, dat1$physical)

# Correlation between mental and temporal
cor(dat1$mental, dat1$temporal)

# Correlation between physical and temporal
cor(dat1$physical, dat1$temporal)
```

## 1b) Summary of Analyses
..

# 2: Summer Youth Development Program
## Data Analyses (SPSS)
See Attached Document (SPSS_02.jpg).

## 2a) Discuss any Validity Issues
...

## 2b) Summary of Analyses
...

# 3: CSSA Predicted Job Performance
## 3a) Data Analyses (R)
### Load & Examine the Dataset
```{r}
# Load the dataset
dat3 <- read.csv('Data/valid.csv', header = T)

# Examine the data
str(dat3)
summary(dat3)

# Compute the correlation between CSSA scores and Performance levels.
cor(dat3$cssa, dat3$perf)

```
The mean of *cssa* and *perf* is 25.1 and 34.06, respectively. Because *r* = 0.57, it appears that there is a positive association between *cssa* and *perf*. As *cssa* increases, *perf* increases as well. 

### Calculate the Prediction

```{r}
# Predict Job Performance using CSSA scores
pred <- lm(perf ~ cssa, data = dat3)
summary(pred)

# Find the r-squared value
summary(pred)$r.squared

# Interpret the regression slope.
pred$coefficients

# Predicted job performance if CSSA score = 10
predict(pred, new=list(cssa = 10))

```
## 3b) Summary of Analyses
..

# 4: Training Program for Caregivers
## 4a) Data Analyses (by hand)
..

## 4b) Calculate Degrees of Freedom
..

## 4c) Summary of Analyses
..