---
title: "Module 05 Assignment"
author: "Alex Schlesener"
date: "2023-09-17"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions

Based on all available data from August 2017, the average Airbnb listing price was $150. You have a simple random sample of active Airbnb listings in New Orleans in August 2018. Compute the sample mean and conduct a test to determine whether the average listing price in New Orleans differs from the average Airbnb listing price in the previous year ($150).Assume  = .05. For this problem, conduct analyses in SPSS and R. Upload an SPSS screenshot and a PDF knitted from R Markdown.

### Import Data

The data file is in the "Data" folder. Use "read.csv" to read in the dataset. 

```{r}

Airbnb <- read.csv("Data/Airbnb_NOLA.csv")

str(Airbnb)
```
### Examine Data
We will now examine the data (summary), and calculate the sample mean and standard deviation of housing prices in 2018.

```{r}

summary(Airbnb)

mean(Airbnb$Price)

sd(Airbnb$Price)

```

### Conduct a Test
Now we will conduct a t-test to compare the sample mean from 2018 ($188.88) with the sample mean from 2017 ($150.00).

```{r}
t.test(Airbnb$Price, mu = 150)
```

## Summary of Results
A one sample t test was performed to evaluate whether there was a significant difference between Airbnb prices in 2018 compared to 2017. The average change in price was $188.88 (*SD* = 229.39). The test was statistically significant, *t*(249) = 2.68, *p* = 0.008. This provides evidence that the prices in Airbnb's went up significantly in 2018, compared to 2017. Cohen's *d* = 0.169.
