---
title: "Module 7 Assignment"
author: "Alex Schlesener"
date: "2023-10-08"
output:
    pdf_document:
      toc: yes
      toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions

You have data from participants who were on one of four types of diets (A, B, C, or D). After three months of dieting, the amount of weight loss was recorded in lb. Determine whether the type of diet was equally effective. Assume $\alpha$ = .05. (a) Conduct relevant analyses by hand to obtain the test statistic. (b) Conduct the appropriate analysis in SPSS and R to confirm the test statistic obtained in part a. (c) Write a brief APA style summary of your Results.

### Import Data

```{r}
Diet <- read.csv("Data/weightloss.csv")
str(Diet)
```

After utilizing the 'str()', we can see that the data frame, "Diet", has 23 rows and 2 columns (variables). The two variables are Weightloss and Diet. Both variables are vectors of integers; however, Diet should be treated as a 'factor'. We will convert 'Diet' to a factor and replace the integer vector to its appropriate code: "1" = "A", "2" = "B", "3" = "C", and "4" = "D".

```{r}
Diet$diet <- factor(Diet$diet, labels = c("A", "B", "C", "D"))
str(Diet)
levels(Diet$diet)
summary(Diet)
```

## Examine the Data

```{r}
# Calculate the Descriptive Statistics by Condition
aggregate(weightloss ~ diet, data = Diet, mean)
aggregate(weightloss ~ diet, data = Diet, sd)

# Plot each mean with the confidence interval using the gplots library.
library(gplots)

plotmeans(weightloss ~ diet, Diet, las = 1,
          ylab = "Weightloss",
          xlab = "Diet",
          n.label = F,
          connect = F)

```

## A) Conduct Releveant Analyses by hand to Obtain the test statistic

### Conduct One-Way Analysis of Variance (By-hand)

```{r}
# Define the sample means of each group
Y1.mean <- 11.0
Y2.mean <- 16.4
Y3.mean <- 18.0
Y4.mean <- 13.2

# Define the N in each group
Y1.N <- 6
Y2.N <- 5
Y3.N <- 7
Y4.N <- 5

# Calculate the grand mean
YG.mean <- (Y1.mean + Y2.mean + Y3.mean + Y4.mean) / 23.0

# Calculate Sum-of-Squares Between
SS.B <- (Y1.N*(Y1.mean - YG.mean)^2) + (Y2.N*(Y2.mean - YG.mean)^2) + (Y3.N*(Y3.mean - YG.mean)^2) + (Y4.N*(Y4.mean - YG.mean)^2)

# Calculate Sum-of-Squares Within
# Group A
SS.W.A <- (12 - Y1.mean)^2 + (10 - Y1.mean)^2 + (13 - Y1.mean)^2 + (9 - Y1.mean)^2 + (13 - Y1.mean)^2 + (9 - Y1.mean)^2

SS.W.B <- (14 - Y2.mean)^2 + (17 - Y2.mean)^2 + (22 - Y2.mean)^2 + (14 - Y2.mean)^2 + (15 - Y2.mean)^2

SS.W.C <- (18 - Y3.mean)^2 + (16 - Y3.mean)^2 + (21 - Y3.mean)^2 + (18 - Y3.mean)^2 + (17 - Y3.mean)^2 + (18 - Y3.mean)^2 + (18 - Y3.mean)^2

SS.W.D <- (12 - Y4.mean)^2 + (11 - Y4.mean)^2 + (13 - Y4.mean)^2 + (16 - Y4.mean)^2 + (14 - Y4.mean)^2

# Total SSQ Within
SS.W <- SS.W.A + SS.W.B + SS.W.C + SS.W.D

# Calculate Sum-of-Squares Total
SS.T <- SS.B + SS.W

# Calculate df Between Groups (alpha - 1)
df.B <- 4 - 1

# Calculate df Within Groups (n - alpha)
df.W <- 23 - 4

# Estimate Variance Between Groups
MS.B <- SS.B / df.B

# Estimate Variance Within Groups
MS.W <- SS.W / df.W 

# Calculate the F Statistic = (Variance between Groups)/(Variance Within Groups)
F.stat <- MS.B / MS.W

```

## B) Conduct the Appropriate Analyses in SPSS and R to confirm the test statistic

### Conduct One-Way Analysis of Variance (SPSS)


### Conduct One-Way Analysis of Variance (R)

```{r}
oneway.diet <- lm(weightloss ~ diet, data=Diet)
summary(oneway.diet)

anova(oneway.diet)

# Check Homogeneity of Variance
library(car)

leveneTest(weightloss ~ diet, data=Diet, center=mean)
```

## C) Write a Brief APA Style Summary of Results
